---
title: "DATA550_final proj"
author: "Jiayi Liu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The **Healthy People 2020 Tobacco Use Objectives** dataset, provided by the CDC, is part of a national initiative aimed at reducing tobacco consumption and its associated health risks in the United States. This report explores trends and key factors influencing progress toward meeting the objectives.

# Data Loading
We load the dataset from a local directory.

```{r load-data}
library(tidyverse)

file_path <- "/Users/liujiayi/DATA550_exercise/final proj/Healthy_People_2020_Tobacco_Use_Objectives_20250128.csv"
data <- read_csv(file_path)

# Display first few rows
head(data)
```

# Exploratory Data Analysis
## Summary Statistics
We first examine the structure and summary of the dataset.

```{r summary-stats}
# Structure of the dataset
str(data)

# Summary statistics
summary(data)
```

## Data Cleaning
We check for missing values and handle them appropriately.

```{r data-cleaning}
# Count missing values
colSums(is.na(data))

data_clean <- data %>% filter(!(is.na(Target) & is.na(Baseline)))

nrow(data_clean) # > 0, proceed to visualization
```

```{r}
# create a direct objective_status column
data_clean <- data_clean %>%
  mutate(Objective_Status = case_when(
    !is.na(Target) & !is.na(Baseline) & Baseline >= Target ~ "Achieved",
    !is.na(Target) & !is.na(Baseline) & Baseline < Target  ~ "In Progress",
    TRUE ~ "Unknown"
  ))

# check if this created meaningful categories
table(data_clean$Objective_Status)
```

# Visualization
## Progress Toward Objectives
A bar plot shows the distribution of progress toward tobacco reduction objectives.

```{r progress-plot}
ggplot(data_clean, aes(x = Objective_Status)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Progress Toward Tobacco Use Objectives",
       x = "Objective Status",
       y = "Count") +
  theme_minimal()
```

**Figure 1**: The distribution of objectives based on their current status.

## Distribution of Baseline Tobacco Use Rates
A histogram showing the distribution of baseline rates.

```{r baseline-distribution}
ggplot(data_clean, aes(x = Baseline)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Baseline Tobacco Use Rates",
       x = "Baseline Rate (%)",
       y = "Count") +
  theme_minimal()
```

## Baseline Tobacco Use Rates by Objective Status
A box plot comparing baseline rates for achieved vs. in-progress objectives.

```{r baseline-boxplot}
ggplot(data_clean, aes(x = Objective_Status, y = Baseline, fill = Objective_Status)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Baseline Tobacco Use Rates by Objective Status",
       x = "Objective Status",
       y = "Baseline Rate (%)") +
  theme_minimal()
```

## Proportion of Objectives by Status
A pie chart displaying the proportion of achieved vs. in-progress objectives.

```{r objective-status-pie}
data_pie <- data_clean %>%
  group_by(Objective_Status) %>%
  summarise(count = n())

ggplot(data_pie, aes(x = "", y = count, fill = Objective_Status)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Proportion of Objectives by Status") +
  theme_void()
```

## Baseline vs. Target Tobacco Use Rates
A scatter plot comparing baseline values to target goals.

```{r baseline-vs-target}
ggplot(data_clean, aes(x = Baseline, y = Target)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Baseline vs. Target Tobacco Use Rates",
       x = "Baseline Rate (%)",
       y = "Target Rate (%)") +
  theme_minimal()
```

# Statistical Analysis
We use a simple linear regression model to assess factors influencing progress.

```{r regression-model}
lm_model <- lm(Baseline ~ Objective_Status, data = data_clean)
summary(lm_model)
```

The regression results provide insights into how objective status influences the baseline rate of tobacco use.

# Conclusion and Recommendations
Based on the findings:

- Most objectives show positive progress, but some lag behind.

- The distribution of baseline rates varies significantly.

- Regression results suggest that objective status significantly impacts progress.

**Recommendations:**
- Strengthen policies targeting lagging objectives.

- Enhance public awareness campaigns.

- Further research to identify barriers to progress.

This analysis provides valuable insights for future tobacco control strategies.
